### 2011-11-04 doc/pldi2012/highlevel.tex
More text for the high level runtime section

\subsubsection{Putting It All Together: Circuit Example}
\label{subsec:cirdependence}

\subsection{Implicit Dataflow Construction}
\label{subsec:dataflow}
In addition to computing the dependences between tasks, we also have to compute where
a task is permitted to acquire its data from when it goes to execute.
It is insufficient to simply record the producer task as there
may be multiple active tasks for a logical region, all of which may produce different physical
instances.  In addition, it is likely that the producer tasks have not been mapped
yet which makes it impossible to know the set of physical instances from which to pull data.

@@

To determine the source and destination abstract instances of a region, we keep track of 
the previous abstract instances that were observed on the way to our target region node
when performing dependence analysis.  These abstract instances describe the set of physical
instances from which valid data for our task can be chosen when it is mapped.  If a region
subtree must be closed then all the abstract instances in that subtree are marked closed
so no additional tasks can register them as sources.  We can also chose to create a new
abstract instance to replace a currently valid abstract instance to avoid a WAR conflict.
By creating an abstract instance we will force the mapper to choose a different physical
instance from the existing ones when mapping is performed.

Abstract instances correspond to a dynamic construction of the dataflow graph.  Each abstract
instance is a node in the dataflow graph corresponding to a specific version of a logical 
region and its instantiations as physical regions.  Tasks are then edges between abstract
instances indicating operations on different versions of data.  By computing this information
during conflict detection, we can record it before tasks begin executing out of order allowing
us to guarantee program correctness in the presence of out of order task execution.  Continuing
the analogy with out-of-order processors, this implicit construction of the dataflow graph
is annalogous to register renaming which is done in order, but allows the processor to remove
WAR dependences and execute instructions in parallel that may have had false dependences.

\subsection{Mapping}
\label{subsec:mapping}

\subsubsection{Placing Tasks}

\subsubsection{Instance Management}
\label{subsec:instmang}

>>>>

\subsubsection{Implicit Dataflow Construction}
\label{subsec:dataflow}
In addition to computing the dependences between tasks, we also have to compute where
a task is permitted to acquire its data from when it goes to execute.
It is insufficient to simply record the producer task as there
may be multiple active tasks for a logical region, all of which may produce different physical
instances.  In addition, it is likely that the producer tasks have not been mapped
yet which makes it impossible to know the set of physical instances from which to pull data.

@@

To determine the source and destination abstract instances of a region, we keep track of 
the previous abstract instances that were observed on the way to our target region node
when performing dependence analysis.  These abstract instances describe the set of physical
instances from which valid data for our task can be chosen when it is mapped.  If a region
subtree must be closed then all the abstract instances in that subtree are marked closed
so no additional tasks can register them as sources.  We can also chose to create a new
abstract instance to replace a currently valid abstract instance to avoid a WAR dependence.
By creating an abstract instance we will force the mapper to choose a different physical
instance from the existing ones when mapping is performed.

Abstract instances correspond to a dynamic construction of the dataflow graph.  Each abstract
instance is a node in the dataflow graph corresponding to a specific version of a logical 
region and its instantiations as physical regions.  Tasks are then edges between abstract
instances indicating operations on different versions of data.  By computing this information
during dependence detection, we can record it before tasks begin executing out of order allowing
us to guarantee program correctness in the presence of out of order task execution.  Continuing
the analogy with out-of-order processors, this implicit construction of the dataflow graph
is annalogous to register renaming which is done in order, but allows the processor to remove
WAR dependences and execute instructions in parallel that may have had false dependences.

\subsubsection{Putting It All Together: Circuit Example}
\label{subsec:cirdependence}
To illustrate concretely how dependence analysis works, we describe the steps for
performing dependence analysis for running circuit example.  When the {\tt simulate\_circuit}
task begins executing there are two region trees rooted at the logical
regions for {\tt all\_nodes} and {\tt all\_wires}.  For conciseness we will only
show the region tree for {\tt all\_nodes}.  At the start of the task, there exists
a single abstract instance at the root node representing the valid set of physical
instances from which sub tasks will be allowed to pull data.  The state of the
system at this point can be observed in figure \ref{fig:circdep1}.

When each of the {\tt compute\_new\_current} tasks are issued they must perform dependence
analysis for two node regions.  In the {\tt private\_node} case they will first open
up the {\tt p\_nodes\_pvs} partition and then open up the {\tt p\_pvt\_nodes} sub partition
for each of their subregions.  Since all of the private regions use a different subregion
of a disjoint partition for the private node accesses, no dependence is detected.  Similarly,
when all the dependence analysis is performed for the {\tt p\_ghost\_nodes} subregions,
each of the necessary partitions are opened.  There is no dependence with any of the private
nodes since all the ghost nodes traverse a different sub region of the disjoint partition
{\tt p\_nodes\_pvs}.  There is also no dependence between any tasks even though the
{\tt p\_ghost\_nodes} partition is aliased because all the ghost accesses are read-only.
The state of the region tree after dependence analysis has been performed for the
{\tt compute\_new\_current} nodes can be seen in figure \ref{fig:circdep2}.

In the second phase of the computation, all of the {\tt distribute\_charge} tasks
will be issued.  For these tasks they will first analyze traverse the region tree
for their subregions containing their private nodes.  The path to these regions are
already open, but when the dependence analysis arrives at the specific private node subregion it will
discover a dependence between the task and the {\tt compute\_new\_currents} task that operated
on that region previously.  This dependence will be recorded and the 
{\tt distribute\_charge} task will be made the new active task as it dominated the
previous task.  For the ghost node regions, the dependence analysis will discover that the
there is a dependence between each of the {\tt distribute\_charge} tasks and all of the
{\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
performing writes and the partition is aliased.  However, the runtime will detect this as
anti-dependences which are resolved by forcing the {\tt distribute\_charge} tasks to
create new valid abstract instances rather than use the same abstract instances as
each of the {\tt compute\_new\_current} tasks.  There are no dependences detected
between each of the {\tt distribute\_charge} tasks on the ghost regions because
the {\tt Simultaneous} coherence property permits multiple tasks to be using the same
partition at the same time even if the subregions they are accessing are aliased.  The
state of the region tree after dependence analysis for {\tt distribute\_charge} can
be seen in figure \ref{fig:circdep3}.

For the final phase of the computation, all of the {\tt update\_voltage} tasks
are issued requiring their private node region and their shared node region.  Again
the private node paths are already open, but a WAR dependence is detected which
requires a new abstract instance to be created to avoid a dependence.  The
shared nodes present a more interesting case.  The first shared node to be registered
will detect a dependence with every prior task since it must open the shared partition
instead of the ghost partition.  A new abstract instance is created at the 
{\tt all\_shared} node and copies are recorded for each valid instance in the ghost partition
back to the new {\tt all\_shared} abstract instance.  After this is complete, the
ghost partition can be opened.  Each corresponding {\tt update\_voltage} task will record
a dependence on the copy operation to close ghost partition as it traverses the {\tt all\_shared}
node ensuring that there is no requirement on the ordering of {\tt update\_voltage}
tasks.  The state of the region tree after this phase can be seen in figure \ref{fig:circdep4}.

\subsection{Mapping}
\label{subsec:mapping}
After tasks have undergone dependence analysis, the tasks are placed on either a ready
queue to be mapped or waiting queue.  A task is placed on the ready queue if all the tasks with
which it registered a dependence have mapped.  If any of the task's dependences have
not mapped then the task is placed on the waiting queue.  The task also registers
itself with the task it depends on so that when the dependent task is mapped it will notify
the task on the waiting queue. Tasks are mapped based on the partial ordering created by 
their data dependences.  By mapping tasks based on data dependences we guarantee the 
semantics of the programming model, while still allowing tasks to execute out of order.

When the scheduler is invoked, it will pull a series of tasks off the ready queue, map them,
and then issue them to be executed by the low level runtime.  The process of mapping a task
consists of several different queries made by the runtime to an application level object
called a {\em mapper}.  A mapper is an object which has control over how both tasks and
data can be placed in the machine.  This idea is a dynamic instantiation of the idea
initially presented in \cite{Fatahalian06} of creating mapping files for specifying how
tasks and data were placed in a memory hierarchy.  Rather than rely on a static mapping
we opted for a dynamic interface that could make decisions both based on application specific
input data as well as the machine specifics both at start-up and throughout the execution
of an application.  The interface for a mapper is described in section \ref{subsec:mapinter}.

\subsubsection{Mapper Interface}
\label{subsec:mapinter}
The interface for a mapper is designed to give the programmer control over how data and tasks
are placed in the machine.  Every mapper must support the following function calls:

\begin{itemize}
\item {\tt rank\_initial\_region\_locations} - When a new logical region is created, the runtime
will ask the mapper to give a ranking of physical memories of where to place the inital instance.

\item {\tt rank\_initial\_partition\_locations} - When a new partition is created, the mapper
must provide a ranking of memory locations for each of the new sub partitions.

\item {\tt select\_initial\_processor} - Given a task to be mapped, select a processor to
send this task to for mapping.

\item {\tt target\_task\_steal} - select a processor to attempt task stealing.  This is explained
in more detail in section \ref{subsec:steal}.

\item {\tt permit\_task\_steal} - also covered in section \ref{subsec:steal}

\item {\tt map\_task\_region} - Given a specific task and the logical region to be mapped, select
a source physical instance for the data, and provide a ranking of memories in which to create
or reuse an existing physical instance.

\item {\tt rank\_copy\_targets} - Given a task and a set of valid instances, generate a set of
rankings for new memories in which to place new phsyical instances.  This operation is used to
manage the copy operations performed when closing a subtree as they allow the programmer to
eagerly migrate regions through the memory hierarchy as copy ups are occuring.

\item {\tt select\_copy\_source} - Given a set of currently valid source instances and destination
physical instance, select the instance from which a copy should be made.
\end{itemize}

These eight task calls give the programmer complete control over where all tasks are executed, and
where all data is placed within the memory hierarchy.  While this interface does give the programmer
complete control over data and task movement, it will never allow the programmer to affect the
correctness of the program.  The highlevel runtime still manages all the task dependences and the
ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
made by any mapper, the resulting program will always be executed correctly.

In order to make it easy for programmers to write programs, we provide an initial implementation 
of the mapper interface that uses simple heuristics to place task and data.  This makes it easy
for the programmer to get an initial program running, while allowing them to write their own
mapper to tune for application or machine specific scenarios at a later time.

Lastly, the runtime supports multiple mappers in use in the same application.  For each task 
call the programmer can choose a mapper to be used for the runtime calls made inside that
task.  This allows for the creation of library code which provides their own mappers indpendent
from any application level mappers that may be created.

\subsubsection{Task Stealing for Load Balance}
\label{subsec:steal}
In order to achieve load balancing in an application the runtime supports an interface
for task stealing.  Our task stealing algorithm is loosely based on the stealing algorithm
used by Cilk\ref{CILK95} with guidance from the mapper about how stealing should be
performed.  Tasks are initially placed on the ready queue on the processor in which they
were registered.  However, a task can be moved to a different processor either by a mapper
explicitly choosing to relocate it, or by a steal request from another processor.

When a processor notices that it has no work on its ready queue, it will inquire into each
mapper asking for a target processor to attempt a steal via the {\tt target\_task\_steal}
call in the mapper interface.  A mapper can either respond with
a target processor, or indicate that no steal should be attempted.  The runtime will then
send steal requests to each processor on behalf of the specific mapper.

When a steal request is received at a processor it will determine the mapper(s) that generated
the list of steal requests.  For each mapper that issued a steal request, the runtime will
give to the local instance of that mapper a list of the tasks in the ready queue that that
specific mapper is responsible for mapping.  The mapper must then determine the set which
it is going to permit to be stolen if any.  Any tasks that the mapper marks as stolen are
sent back to the stealing processor.  By only allowing mappers to make steal requests about
the tasks they own, we are giving the mappers complete control over where tasks are mapped.
In addition, we are also giving the mappers control over how steal requests are made for
the case where the mapper has better knowledge about where additional work is than the 
runtime.  Finally, note that this mapping interface is sufficiently powerful to allow a user to
disable all stealing if desired, and to map tasks to explicit processors with no interference
from the runtime.
